---
title: "Coursera 8 Assignment"
author: "JSoh"
date: "Saturday, August 22, 2015"
output: html_document
---

```{r setoptions,echo=FALSE,message=FALSE,warning=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
```
##Overview

The goal of this assignment is to predict the manner in which people exercised using data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 
The aim is to build a prediction model to predict the "classe" variable in the data set. 

##Data download and processing
Training and test data is downloaded and loaded into respective data frames.
```{r loadData,cache=TRUE}
# Download trainig and test data file from source
urlToDownloadTrainData ="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlToDownloadTestData ="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

if(!file.exists("pml-training.csv")){
    download.file(url = urlToDownloadTrainData, destfile = "pml-training.csv",method = "curl")
}
if(!file.exists("pml-test.csv")){
    download.file(url = urlToDownloadTestData, destfile = "pml-test.csv", method="curl")
}
# Read and load the data file into a dataframe
traindata <- read.csv(file="pml-training.csv",header=TRUE,na.strings = c(""," ",NA,"#DIV/0!"))
testdata <- read.csv(file="pml-testing.csv",header=TRUE,na.strings = c(""," ",NA,"#DIV/0!"))
```

After inspecting the data, we clean both the training and test data sets by removing columns that has at least one NA value as well as the first 5 columns as the data is not required:
```{r cleandata, cache=TRUE, results='hide'}
head(traindata)
naCols <- which(unname(sapply(traindata,anyNA)))
cleanTrainData <- traindata[,-c(1:5,naCols)]
cleanTestData <- testdata[,-c(1:5,naCols)]
```

##Training the data
Proceed with partitioning part of the training data set for training the prediction mdoel and the test set for validating the prediction model. 
```{r train, cache=TRUE, results='hide'}
library(caret)
train <- createDataPartition(cleanTrainData$classe,p=0.7,list=FALSE)
trainset <- cleanTrainData[train,]
testTrainset <- cleanTrainData[-train,]
```
We will use random forest machine learning model to formulate the prediction model:
```{r RFtraining,cache=TRUE}
library(randomForest)
set.seed(1234)
rfmod <- randomForest(classe~.,data=trainset)
rfmod
```
Looking at the OOB error, it seems that the error rate is pretty low at 0.28%.
To assess the accuracy of the prediction model we will use the test data set that was carved out from the original training data set:
```{r accuracy,cache=TRUE}
accuracy <- confusionMatrix(predict(rfmod,testTrainset[,-55]),testTrainset$classe)
accuracy$overall
errFunc = function(values, prediction) {
    sum(prediction != values)/length(values)
}
oosamplerr <- errFunc(testTrainset$classe,predict(rfmod,testTrainset[,-55]))
oosamplerr
```